---
title: "Cloud Storage Access for GDS Files"
author: "Xiuwen Zheng"
date: "2026-05-01"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{Cloud Storage Access for GDS Files}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE)
```


# Introduction

GDS (Genomic Data Structure) is a high-performance file format for storing and accessing large-scale genomic data, implemented by the `r BiocStyle::Biocpkg("gdsfmt")` package. It supports hierarchical data organization with efficient random access and data compression.

The `r BiocStyle::Biocpkg("gdscloud")` package extends `r BiocStyle::Biocpkg("gdsfmt")` to provide transparent read-only access to GDS files stored on cloud storage services. It uses HTTP Range requests via libcurl with an efficient LRU block cache to minimize network overhead.

Supported backends:

- **HTTP/HTTPS** — any `http://` or `https://` URL (public or authenticated)
- **Amazon S3** — URLs of the form `s3://bucket/key`
- **Google Cloud Storage** — URLs of the form `gs://bucket/key`
- **Azure Blob Storage** — URLs of the form `az://container/blob`


# Installation

```R
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("gdscloud")
```


# Supported URL Schemes

gdscloud recognizes the following URL schemes:

```{r schemes}
library(gdscloud)
gdsCloudSchemes()
```


# Quick Start

Once gdscloud is loaded, `openfn.gds()` from the gdsfmt package automatically recognizes cloud URLs and opens them transparently:

```{r quickstart}
library(gdscloud)

# Open a GDS file from S3 — transparent via openfn.gds()
gds <- openfn.gds("s3://gds-stat/download/hapmap/hapmap_r23a.gds")
gds

table(read.gdsn(index.gdsn(gds, "chromosome")))

summary(read.gdsn(index.gdsn(gds, "position")))

closefn.gds(gds)
```

Alternatively, call `gdsCloudOpen()` directly with a cloud URL (see `?gdsCloudOpen`); the S3, GCS, Azure, and HTTP examples below use it explicitly.


# Authentication

Each cloud backend requires credentials. Credentials can be set either via environment variables (recommended for non-interactive use) or via R functions (convenient for interactive sessions). R-configured credentials take priority over environment variables.

The secrets shown below are placeholders. Avoid typing real credentials directly into R scripts or on a shell command line, since both are saved to the R (`.Rhistory`) and shell histories. To keep secrets out of your history, store them in an OS credential store with the `r BiocStyle::CRANpkg("keyring")` package and retrieve them at run time:

```{r keyring, eval=FALSE}
# one-time, interactive: store the secret in the OS keychain
keyring::key_set("gdscloud", "aws_secret_access_key")

# later, in a script: retrieve without exposing it in history
gdsCloudConfigS3(
    aws_access_key_id     = keyring::key_get("gdscloud", "aws_access_key_id"),
    aws_secret_access_key = keyring::key_get("gdscloud", "aws_secret_access_key"),
    region                = "us-east-1"
)
```

## HTTP/HTTPS

For public URLs (e.g., files hosted on a web server with Range request support), no credentials are needed:

```{r open-http-public, eval=FALSE}
gds <- gdsCloudOpen("https://example.com/path/to/file.gds")
closefn.gds(gds)
```

For authenticated HTTP endpoints, set a Bearer token:

**Environment variable:**

```bash
export GDSCLOUD_HTTP_TOKEN=your_bearer_token
```

**R configuration:**

```{r auth-http}
gdsCloudConfigHTTP(bearer_token = "your_bearer_token")

# URL-specific token for a private server
gdsCloudConfigHTTP(
    bearer_token = "token_for_private_server",
    url = "https://private.example.com/"
)
```

Then open:

```{r open-http, eval=FALSE}
gds <- gdsCloudOpen("https://private.example.com/data/file.gds")
closefn.gds(gds)
```

## Amazon S3

**Environment variables:**

```bash
export AWS_ACCESS_KEY_ID=your_key
export AWS_SECRET_ACCESS_KEY=your_secret
export AWS_DEFAULT_REGION=us-east-1
# Optional for temporary credentials:
export AWS_SESSION_TOKEN=your_token
```

**R configuration:**

```{r auth-s3}
gdsCloudConfigS3(
    aws_access_key_id = "your_key",
    aws_secret_access_key = "your_secret",
    region = "us-east-1"
)
```

Then open:

```{r open-s3, eval=FALSE}
gds <- gdsCloudOpen("s3://my-bucket/path/to/file.gds")
# ... work with the file ...
closefn.gds(gds)
```

## Google Cloud Storage

**Environment variable:**

```bash
export GCS_ACCESS_TOKEN=your_oauth2_token
```

**R configuration:**

```{r auth-gcs}
gdsCloudConfigGCS(access_token = "your_oauth2_token")
```

Then open:

```{r open-gcs, eval=FALSE}
gds <- gdsCloudOpen("gs://my-bucket/path/to/file.gds")
closefn.gds(gds)
```

## Azure Blob Storage

**Environment variables:**

```bash
export AZURE_STORAGE_ACCOUNT=your_account
export AZURE_STORAGE_KEY=your_key
# Or use a SAS token instead:
export AZURE_STORAGE_SAS_TOKEN=your_sas_token
```

**R configuration:**

```{r auth-azure}
gdsCloudConfigAzure(
    account_name = "mystorageaccount",
    account_key = "base64encodedkey=="
)
# Or with SAS token:
gdsCloudConfigAzure(
    account_name = "mystorageaccount",
    sas_token = "sv=2021-06-08&ss=b&srt=co&sp=r..."
)
```

Then open:

```{r open-azure, eval=FALSE}
gds <- gdsCloudOpen("az://my-container/path/to/file.gds")
closefn.gds(gds)
```

## URL-specific credentials

Sometimes different URLs need different credentials — for example, two S3 buckets owned by different accounts, or a mix of public and private resources. Each `gdsCloudConfig*()` function accepts an optional `url` argument that associates the supplied credentials with a URL prefix rather than the global defaults:

```{r auth-url}
# Different keys for two S3 buckets
gdsCloudConfigS3(
    aws_access_key_id = "KEY_A",
    aws_secret_access_key = "SECRET_A",
    url = "s3://bucket-a/"
)
gdsCloudConfigS3(
    aws_access_key_id = "KEY_B",
    aws_secret_access_key = "SECRET_B",
    url = "s3://bucket-b/"
)

# You can also scope credentials to a sub-prefix within a bucket
gdsCloudConfigS3(
    aws_access_key_id = "KEY_SHARED",
    aws_secret_access_key = "SECRET_SHARED",
    url = "s3://bucket-a/shared/"
)

# Remove a previously registered URL-specific entry
gdsCloudConfigS3(url = "s3://bucket-a/")
```

When opening a URL, credentials are resolved in the following order (the first non-empty value wins for each field):

1. the URL-specific entry whose registered prefix is the longest prefix of the opened URL (within the same scheme);
2. the global values set by `gdsCloudConfig*()` with `url=NULL`;
3. the corresponding environment variable.

The URL scheme passed via `url=` must match the function — `http://` or `https://` for `gdsCloudConfigHTTP()`, `s3://` for `gdsCloudConfigS3()`, `gs://` for `gdsCloudConfigGCS()`, and `az://` for `gdsCloudConfigAzure()`. Registered prefixes are normalized by appending a trailing `/` if missing, so `"s3://bucket"` and `"s3://bucket/"` behave identically.


# Cache Control

Cloud access uses an LRU block cache (1 MB blocks) to minimize HTTP requests. The default cache size is 64 MB per stream.

```{r cache}
# Set the default cache size for new streams (in MB)
gdsCloudCacheSize(128)

# Clear all cached data
gdsCloudCacheClear()

# Display cache statistics
gdsCloudCacheInfo()

# List all open cloud streams
gdsCloudList()
```

Increasing cache size is beneficial when working with large files that require many random seeks (e.g., subsetting genotype matrices by sample and variant).


# Integration with SeqArray

Since gdscloud works transparently through `openfn.gds()`, packages built on gdsfmt, such as `r BiocStyle::Biocpkg("SeqArray")`, can open cloud-hosted files directly.

**Note:** SeqArray >= v1.53.1 is recommended for full cloud support. This version allows `seqParallel()` to automatically load cloud-related packages on worker processes, so parallel operations on cloud-hosted GDS files work seamlessly.

```{r seqarray, eval=FALSE}
library(SeqArray)
library(gdscloud)

# Open a SeqArray GDS file from S3
gds <- seqOpen("s3://gds-stat/download/1000g/2022/1kGP_high_coverage_Illumina.allchr.filtered.SNV_INDEL_SV_phased_panel.gds")
gds
## File: s3://gds-stat/download/1000g/2022/1kGP_high_coverage_Illumina.allchr.filtered.SNV_INDEL_SV_phased_panel.gds (2.4G)
## +    [  ] *
## |--+ description   [  ] *
## |--+ sample.id   { Str8 3202 LZMA_ra(6.61%), 1.7K } *
## |--+ variant.id   { Int32 73554796 LZMA_ra(2.36%), 6.6M } *
## |--+ position   { Int32 73554796 LZMA_ra(27.9%), 78.4M } *
## |--+ chromosome   { Str8 73554796 LZMA_ra(0.01%), 25.3K } *
## |--+ allele   { Str8 73554796 LZMA_ra(15.8%), 51.4M } *
## |--+ genotype   [  ] *
## |  |--+ data   { Bit2 2x3202x73554796 LZMA_ra(1.77%), 1.9G } *
## |  |--+ extra.index   { Int32 3x0 LZMA_ra, 18B } *
## |  \--+ extra   { Int16 0 LZMA_ra, 18B }
## |--+ phase   [  ]
## |  |--+ data   { Bit1 3202x73554796 LZMA_ra(0.01%), 4.1M } *
## |  |--+ extra.index   { Int32 3x0 LZMA_ra, 18B } *
## |  \--+ extra   { Bit1 0 LZMA_ra, 18B }
## |--+ annotation   [  ]
## |  |--+ id   { Str8 73554796 LZMA_ra(17.2%), 186.3M } *
## |  |--+ qual   { Float32 73554796 LZMA_ra(0.01%), 42.0K } *
## |  |--+ filter   { Int32,factor 73554796 LZMA_ra(0.01%), 42.0K } *
## |  |--+ info   [  ]
## |  |  |--+ AF   { Float32 73554796 LZMA_ra(23.6%), 66.3M } *
## |  |  |--+ AC   { Int32 73554796 LZMA_ra(22.3%), 62.5M } *
## |  |  |--+ CM   { Float32 73554796 LZMA_ra(6.08%), 17.1M } *
## |  |  |--+ AN   { Int32 73554796 LZMA_ra(0.01%), 42.0K } *
## |  |  \--+ SVTYPE   { Str8 73554796 LZMA_ra(0.33%), 240.0K } *
## |  \--+ format   [  ]
## \--+ sample.annotation   [  ]

seqSummary(gds)
## File: s3://gds-stat/download/1000g/2022/1kGP_high_coverage_Illumina.allchr.filtered.SNV_INDEL_SV_phased_panel.gds
## Format Version: v1.0
## Reference: unknown
## Ploidy: 2
## Number of samples: 3,202
## Number of variants: 73,554,796
## Chromosomes:
##     chr1 : 5759060, chr2 : 6088598, chr3 : 4983185, chr4 : 4875465, chr5 : 4536819, chr6 : 4315217
##     chr7 : 4137254, chr8 : 3886222, chr9 : 3165513, chr10: 3495473, chr11: 3423341, chr12: 3332788
##     chr13: 2509179, chr14: 2290400, chr15: 2109285, chr16: 2362361, chr17: 2073624, chr18: 1963845
##     chr19: 1670692, chr20: 1644384, chr21: 1002753, chr22: 1066557, chrX : 2862781
## ...

seqClose(gds)
```

No code changes are needed in downstream packages; loading gdscloud is sufficient to enable cloud URL support.


# Session Information

```{r pkg, echo=FALSE}
library(gdscloud, quietly=TRUE)
```

```{r session}
sessionInfo()
```
