---
title: "Cloud Storage Access for GDS Files"
author: "Xiuwen Zheng"
date: "2026-05-01"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{Cloud Storage Access for GDS Files}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```


# Introduction

GDS (Genomic Data Structure) is a high-performance file format for storing and accessing large-scale genomic data, implemented by the [gdsfmt](https://bioconductor.org/packages/gdsfmt) package. It supports hierarchical data organization with efficient random access and data compression.

The **gdscloud** package extends gdsfmt to provide transparent read-only access to GDS files stored on cloud storage services. It uses HTTP Range requests via libcurl with an efficient LRU block cache to minimize network overhead.

Supported cloud storage backends:

- **Amazon S3** — URLs of the form `s3://bucket/key`
- **Google Cloud Storage** — URLs of the form `gs://bucket/key`
- **Azure Blob Storage** — URLs of the form `az://container/blob`


# Installation

```{r install}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("gdscloud")
```


# Quick Start

Once gdscloud is loaded, `openfn.gds()` from the gdsfmt package automatically recognizes cloud URLs and opens them transparently:

```{r quickstart}
library(gdscloud)

# Open a GDS file from S3 — transparent via openfn.gds()
gds <- openfn.gds("s3://gds-stat/download/hapmap/hapmap_r23a.gds")
gds
## File: s3://gds-stat/download/hapmap/hapmap_r23a.gds (86.5M)
## +    [  ] *
## |--+ description   [  ] *
## |--+ sample.id   { Str8 270 LZMA_ra(17.3%), 381B } *
## |--+ variant.id   { Int32 4098136 LZMA_ra(3.19%), 511.4K } *
## |--+ position   { Int32 4098136 LZMA_ra(48.2%), 7.5M } *
## |--+ chromosome   { Str8 4098136 LZMA_ra(0.02%), 1.7K } *
## |--+ allele   { Str8 4098136 LZMA_ra(12.8%), 2.0M } *
## |--+ genotype   [  ] *
## |  |--+ data   { Bit2 2x270x4098136 LZMA_ra(12.4%), 65.4M } *
## |  |--+ extra.index   { Int32 3x0 LZMA_ra, 18B } *
## |  \--+ extra   { Int16 0 LZMA_ra, 18B }
## |--+ phase   [  ]
## |  |--+ data   { Bit1 270x4098136 LZMA_ra(0.01%), 19.8K } *
## |  |--+ extra.index   { Int32 3x0 LZMA_ra, 18B } *
## |  \--+ extra   { Bit1 0 LZMA_ra, 18B }
## |--+ annotation   [  ]
## |  |--+ id   { Str8 4098136 LZMA_ra(27.7%), 11.1M } *
## |  |--+ qual   { Float32 4098136 LZMA_ra(0.02%), 2.5K } *
## |  |--+ filter   { Int32,factor 4098136 LZMA_ra(0.02%), 2.5K } *
## |  |--+ info   [  ]
## |  \--+ format   [  ]
## \--+ sample.annotation   [  ]
##    |--+ family   { Str8 270 LZMA_ra(17.3%), 381B } *
##    |--+ father   { Str8 270 LZMA_ra(28.2%), 261B } *
##    |--+ mother   { Str8 270 LZMA_ra(28.2%), 261B } *
##    |--+ sex   { Str8 270 LZMA_ra(27.0%), 153B } *
##    \--+ phenotype   { Int32 270 LZMA_ra(8.70%), 101B } *

read.gdsn(index.gdsn(gds, "chromosome"))
## "1", ...

read.gdsn(index.gdsn(gds, "position"))
## 45162 45257 45413 46844 72434 72515 ...

closefn.gds(gds)
```

Alternatively, use the explicit function:

```{r explicit}
gds <- gdsCloudOpen("s3://my-bucket/data/example.gds")
closefn.gds(gds)
```


# Authentication

Each cloud backend requires credentials. Credentials can be set either via environment variables (recommended for non-interactive use) or via R functions (convenient for interactive sessions). R-configured credentials take priority over environment variables.

## Amazon S3

**Environment variables:**

```bash
export AWS_ACCESS_KEY_ID=your_key
export AWS_SECRET_ACCESS_KEY=your_secret
export AWS_DEFAULT_REGION=us-east-1
# Optional for temporary credentials:
export AWS_SESSION_TOKEN=your_token
```

**R configuration:**

```{r auth-s3}
gdsCloudConfigS3(
    aws_access_key_id = "your_key",
    aws_secret_access_key = "your_secret",
    region = "us-east-1"
)
```

Then open:

```{r open-s3}
gds <- gdsCloudOpen("s3://my-bucket/path/to/file.gds")
# ... work with the file ...
closefn.gds(gds)
```

## Google Cloud Storage

**Environment variable:**

```bash
export GCS_ACCESS_TOKEN=your_oauth2_token
```

**R configuration:**

```{r auth-gcs}
gdsCloudConfigGCS(access_token = "your_oauth2_token")
```

Then open:

```{r open-gcs}
gds <- gdsCloudOpen("gs://my-bucket/path/to/file.gds")
closefn.gds(gds)
```

## Azure Blob Storage

**Environment variables:**

```bash
export AZURE_STORAGE_ACCOUNT=your_account
export AZURE_STORAGE_KEY=your_key
# Or use a SAS token instead:
export AZURE_STORAGE_SAS_TOKEN=your_sas_token
```

**R configuration:**

```{r auth-azure}
gdsCloudConfigAzure(
    account_name = "mystorageaccount",
    account_key = "base64encodedkey=="
)
# Or with SAS token:
gdsCloudConfigAzure(
    account_name = "mystorageaccount",
    sas_token = "sv=2021-06-08&ss=b&srt=co&sp=r..."
)
```

Then open:

```{r open-azure}
gds <- gdsCloudOpen("az://my-container/path/to/file.gds")
closefn.gds(gds)
```


# Cache Control

Cloud access uses an LRU block cache (1 MB blocks) to minimize HTTP requests. The default cache size is 64 MB per stream.

```{r cache}
# Set the default cache size for new streams (in MB)
gdsCloudCacheSize(128)

# Clear all cached data
gdsCloudCacheClear()

# Display cache statistics
gdsCloudInfo()

# List all open cloud streams
gdsCloudList()
```

Increasing cache size is beneficial when working with large files that require many random seeks (e.g., subsetting genotype matrices by sample and variant).


# Integration with SeqArray

Since gdscloud works transparently through `openfn.gds()`, packages built on gdsfmt, such as [SeqArray](https://bioconductor.org/packages/SeqArray), can open cloud-hosted files directly.

**Note:** SeqArray >= v1.53.1 is recommended for full cloud support. This version allows `seqParallel()` to automatically load cloud-related packages on worker processes, so parallel operations on cloud-hosted GDS files work seamlessly.

```{r seqarray}
library(SeqArray)
library(gdscloud)

# Open a SeqArray GDS file from S3
gds <- seqOpen("s3://gds-stat/download/1000g/2022/1kGP_high_coverage_Illumina.allchr.filtered.SNV_INDEL_SV_phased_panel.gds")
gds
## File: s3://gds-stat/download/1000g/2022/1kGP_high_coverage_Illumina.allchr.filtered.SNV_INDEL_SV_phased_panel.gds (2.4G)
## +    [  ] *
## |--+ description   [  ] *
## |--+ sample.id   { Str8 3202 LZMA_ra(6.61%), 1.7K } *
## |--+ variant.id   { Int32 73554796 LZMA_ra(2.36%), 6.6M } *
## |--+ position   { Int32 73554796 LZMA_ra(27.9%), 78.4M } *
## |--+ chromosome   { Str8 73554796 LZMA_ra(0.01%), 25.3K } *
## |--+ allele   { Str8 73554796 LZMA_ra(15.8%), 51.4M } *
## |--+ genotype   [  ] *
## |  |--+ data   { Bit2 2x3202x73554796 LZMA_ra(1.77%), 1.9G } *
## |  |--+ extra.index   { Int32 3x0 LZMA_ra, 18B } *
## |  \--+ extra   { Int16 0 LZMA_ra, 18B }
## |--+ phase   [  ]
## |  |--+ data   { Bit1 3202x73554796 LZMA_ra(0.01%), 4.1M } *
## |  |--+ extra.index   { Int32 3x0 LZMA_ra, 18B } *
## |  \--+ extra   { Bit1 0 LZMA_ra, 18B }
## |--+ annotation   [  ]
## |  |--+ id   { Str8 73554796 LZMA_ra(17.2%), 186.3M } *
## |  |--+ qual   { Float32 73554796 LZMA_ra(0.01%), 42.0K } *
## |  |--+ filter   { Int32,factor 73554796 LZMA_ra(0.01%), 42.0K } *
## |  |--+ info   [  ]
## |  |  |--+ AF   { Float32 73554796 LZMA_ra(23.6%), 66.3M } *
## |  |  |--+ AC   { Int32 73554796 LZMA_ra(22.3%), 62.5M } *
## |  |  |--+ CM   { Float32 73554796 LZMA_ra(6.08%), 17.1M } *
## |  |  |--+ AN   { Int32 73554796 LZMA_ra(0.01%), 42.0K } *
## |  |  \--+ SVTYPE   { Str8 73554796 LZMA_ra(0.33%), 240.0K } *
## |  \--+ format   [  ]
## \--+ sample.annotation   [  ]

seqSummary(gds)
## File: s3://gds-stat/download/1000g/2022/1kGP_high_coverage_Illumina.allchr.filtered.SNV_INDEL_SV_phased_panel.gds
## Format Version: v1.0
## Reference: unknown
## Ploidy: 2
## Number of samples: 3,202
## Number of variants: 73,554,796
## Chromosomes:
##     chr1 : 5759060, chr2 : 6088598, chr3 : 4983185, chr4 : 4875465, chr5 : 4536819, chr6 : 4315217
##     chr7 : 4137254, chr8 : 3886222, chr9 : 3165513, chr10: 3495473, chr11: 3423341, chr12: 3332788
##     chr13: 2509179, chr14: 2290400, chr15: 2109285, chr16: 2362361, chr17: 2073624, chr18: 1963845
##     chr19: 1670692, chr20: 1644384, chr21: 1002753, chr22: 1066557, chrX : 2862781
## ...

seqClose(gds)
```

No code changes are needed in downstream packages; loading gdscloud is sufficient to enable cloud URL support.


# Session Information

```{r pkg, eval=TRUE, echo=FALSE}
library(gdscloud, quietly=TRUE)
```

```{r session, eval=TRUE}
sessionInfo()
```
